---
title: "Lesson 2 - ggplot2"
subtitle: "Introduction to Data Visualization in R"
author: "JHU Data Services"
date: "2021-11-03"
output:
  html_document: 
    fig_caption: false
    css: ".\/config\/style.css"
    toc: true
    toc_float: true
    toc_depth: 3
  pdf_document: default
fig_width: 6 
fig_height: 6
---

# Introduction to ggplot2

---

ggplot2 is an R package for producing graphics, and it is particularly well suited to producing statistical graphics. ggplot2 is based on the Grammar of Graphics, which describes a series of components that make up graphs. ggplot2 allows you to build graphs out of these independent components, much like Legos. Rather than be limited to a set number of predefined graphs, you can create a vast variety of graphics by adding, removing, and mixing these components.

Today we will learn what these components are, and how we can combine them to create effective graphics in R.

## Key components

Each ggplot2 plot has three key components:

* **Data**: The dataframe you would like to plot.
* **Aesthetic mappings** between the variables of interest in your data, and a visual property (like size, color, or position)
* At least one **layer**. Layers describe how observations should be rendered (bars, points, etc) and are created with the **geom** function.

# Lets Practice!

---

## Load the libraries and data.

---

Let's load in the Palmer penguin data, by importing the `palmerpenguins` package. If you have not already installed this package, you can do so now by running `install.packages('palmerpenguins')` in your R console. We will also load our libraries for plotting.
```{r setup, eval=TRUE}

# load libraries
library(ggplot2)
library(scales)
library(RColorBrewer)
library(palmerpenguins)

# load in dataset
data(package = "palmerpenguins")
```

## Explore the data

--- 

Let's explore the contents of the data. We can use the `head` command to look at the first few rows of the data. 
```{r explore_data, eval=TRUE}
head(penguins)
```

## Using ggplot2

--- 

### Plotting Data

--- 

When we create a plot with `ggplot2()` we do it incrementally, by adding a layer at a time. Lets begin by calling `ggplot()` and passing our dataset (`penguin`).

::: activity

#### Activity

What happens when we run: `ggplot(data = penguins)`?

:::

```{r ggplot2_skeleton, eval=TRUE}
ggplot(data = penguins)
```

Nothing is produced! This is because we need to add a layer. Let's add a `geom` layer, which will specify the geometric object we want to use to draw each observation. Let's use `geom_point()` to create a scatter plot. Try to run the code again with `geom_point()` added and see what happens?


```{r ggplot2_points, eval=FALSE}
ggplot(data = penguins) +
  geom_point()

```

We received an error because we haven't specified what elements of our data we want to plot.

Lets map to variables in our data to aesthetic elements. In this case let's map `body_mass_g` to the x-axis, and `bill_length_mm` to the y-axis. 

```{r ggplot2_points_aesth, eval=TRUE}
ggplot(data = penguins) +
  geom_point(aes(x =body_mass_g, 
                 y = bill_length_mm))
```
It worked! 

### Modifying Aesthetics

---

Now lets configure another aesthetic of `geom_point()` by specifying a color for the points, using the argument `color`. Lets use the color `"steelblue"`.

:::activity
#### Activity

Add `color = "steelblue"` as an additional argument to `aes()` within geom_point

What happens?
:::

```{r ggplot2_points_aes_color, eval=TRUE}

ggplot(data = penguins) +
  geom_point(aes(x = body_mass_g, 
                 y = bill_length_mm,
                 color = "steelblue")) 
```

The points are colored red, even though we explicitly specified that they be *steelblue*. This unexpected behaviour is related to how ggplot2 handles aesthetic arguments. Those that are defined within `aes()` map data to an aesthetic using a **scale**.

Those that are defined outside of `aes()` are set to a fixed value. Let's try again, but move `color = "steelblue"` outside of `aes()`.

```{r ggplot2_points_color, eval=TRUE}
ggplot(data = penguins) +
  geom_point(aes(x = body_mass_g, 
                 y = bill_length_mm),
                 color = "steelblue") 
```

That produces the result we expected. Remember, if you want to set an aesthetic to a fixed value (such as setting points to the color steelblue) then do it **outside** of `aes()`. If you want to map some variable of your data to an aesthetic automatically, then do it **inside** `aes()`

Let's try mapping the species variable to color:

```{r ggplot2_points_color_aes, eval=TRUE}
ggplot(data = penguins) +
  geom_point(aes(x = body_mass_g, 
                 y = bill_length_mm,
                 color = species))
```

This produces points colored by species, and automatically creates an appropriately named legend. The individual points look a little small, so let's increase their size.

```{r ggplot2_points_size, eval=TRUE}
ggplot(data = penguins) +
  geom_point(aes(x = body_mass_g, 
                 y = bill_length_mm,
                 color = species),
             size = 3)
```

Now let's add transparency so that we can see points being occluded by other points.

```{r ggplot2_points_alpha, eval=TRUE}
ggplot(data = penguins) +
  geom_point(aes(x = body_mass_g, 
                 y = bill_length_mm,
                 color = species),
             size = 3,
             alpha = .5)
```

### Using RColorBrewer Palettes

---

This is looking better, but the default use of red and green is not colorblind friendly. Let's correct this by using a palette from RColorBrewer. Fortunately, with ggplot2 we can pass the palette directly in to ggplot2:

```{r ggplot2_points_rcolorbrewer, eval=TRUE}
ggplot(data = penguins) +
  geom_point(aes(x = body_mass_g, 
                 y = bill_length_mm,
                 color = species),
             size = 3,
             alpha = .5) +
  scale_color_brewer(palette = "Dark2")
```

Now let's clean up those axis labels:

```{r ggplot2_points_labels, eval=TRUE}
ggplot(data = penguins) +
  geom_point(aes(x = body_mass_g, 
                 y = bill_length_mm,
                 color = species),
             size = 3,
             alpha = .5) +
  scale_color_brewer(palette = "Dark2") +
  xlab("Body Mass (g)") +
  ylab("Bill Length (mm)")
```

### Themes

---

Remember, never trust the defaults. Although ggplot2 plots look much nicer than base R, that does not mean they are ready for publication. There are a few things we can do to clean this plot up. Let's start my removing that unnecessary grey background. We can do this easily by applying the **minimal** theme:
 
```{r ggplot2_theme, eval=TRUE}
ggplot(data = penguins) +
  geom_point(aes(x = body_mass_g, 
                 y = bill_length_mm,
                 color = species),
             size = 3,
             alpha = .5) +
  scale_color_brewer(palette = "Dark2") +
  xlab("Body Mass (g)") +
  ylab("Bill Length (mm)") +
  theme_minimal()
```
 
Lets increase the size of our x and y label font and axis label font to make the labels more visible. To do this, we will need to manually modify the theme:

```{r ggplot2_manual_theme, eval=TRUE}
ggplot(data = penguins) +
  geom_point(aes(x = body_mass_g, 
                 y = bill_length_mm,
                 color = species),
             size = 3,
             alpha = .5) +
  scale_color_brewer(palette = "Dark2") +
  xlab("Body Mass (g)") +
  ylab("Bill Length (mm)") +
  theme_minimal() +
  theme(axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        axis.text = element_text(size = 12)) 
```

:::activity

#### Activity

What happens when you switch the positions of `theme()` and `theme_minimal()`?
:::

```{r ggplot2_manual_theme_swap, eval=TRUE}
ggplot(data = penguins) +
  geom_point(aes(x = body_mass_g, 
                 y = bill_length_mm,
                 color = species),
             size = 3,
             alpha = .5) +
  scale_color_brewer(palette = "Dark2") +
  xlab("Body Mass (g)") +
  ylab("Bill Length (mm)") +
  theme(axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        axis.text = element_text(size = 12)) +
  theme_minimal() 
```

All of our modifications are lost. This is because the order of our theme matters. Because we call `theme_minimal()` after `theme()`, we overwrite any changes we made with the defaults from `theme_minimal()`. If we want to modify how `theme_minimal()` looks we do it after calling `theme_minimal()`.