---
title: "Lesson 2 - ggplot2"
subtitle: "Introduction to Data Visualization in R"
author: "JHU Data Services"
date: "2021-11-03"
output:
  html_document: 
    fig_caption: false
    css: ".\/config\/style.css"
    toc: true
    toc_float: true
    toc_depth: 3
  pdf_document: default
fig_width: 6 
fig_height: 6
---

### Load the libraries and data.

---

Let's load in the Palmer penguin data, by importing the `palmerpenguins` package. If you have not already installed this package, you can do so now by running `install.packages('palmerpenguins')` in your R console. We will also load our libraries for plotting.
```{r setup, eval=TRUE}

# load libraries
library(ggplot2)
library(scales)
library(RColorBrewer)
library(palmerpenguins)

# load in dataset
data(package = "palmerpenguins")
```

### Explore the data

--- 

Let's explore the contents of the data. We can use the `head` command to look at the first few rows of the data. 
```{r explore_data, eval=TRUE}
head(penguins)
```

### Plotting in ggplot2

When we create a plot with `ggplot2()` we do it incrementally, by adding a layer at a time. Lets begin by calling `ggplot()` and passing our dataset (`penguin`), and specifying the aesthetic mappings we would like to visualize (body mass and bill length).
```{r ggplot2_skeleton, eval=TRUE}
ggplot(data = penguins, aes(x = body_mass_g, 
                            y = bill_length_mm))
```

Nothing is produced! This is bacause we need to add a layer. Let's add a `geom` layer, which will specify the geometric object we want to use to draw each observation.

```{r ggplot2_points, eval=TRUE}
ggplot(data = penguins, aes(x = body_mass_g, 
                            y = bill_length_mm)) +
  geom_point()

```

